---
title: "Annotated DPI clusters in BED format."
output:
  html_document: 
    keep_md: yes
  html_notebook: default
---

# Annotated DPI clusters in BED format.

[FANTOM5](http://fantom.gsc.riken.jp/5/) CAGE peaks
([DPI clusters](http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/))
are avaliable as BED files for their coordinates, supplemented by an annotation
file.  This annotation files provides a short description that contains the
gene symbol when possible.  Here, I replace the "name" field of the BED file
by the "short_description" field of the annotation file.  I do it in R because
the files are not sorted in the same order, and I am more comfortable
with join operations in R than in shell.

## Get and check the data in the command line:

Download the human DPI clusters from the FANTOM5 website.

```{sh}
curl --silent http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/hg19.cage_peak_phase1and2combined_ann.txt.gz | gunzip > hg19.cage_peak_phase1and2combined_ann.txt
curl --silent http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/hg19.cage_peak_phase1and2combined_coord.bed.gz | gunzip > hg19.cage_peak_phase1and2combined_coord.bed
```

The file `hg19.cage_peak_phase1and2combined_ann.txt` contains annotations.

```{sh}
head hg19.cage_peak_phase1and2combined_ann.txt
```

The files `hg19.cage_peak_phase1and2combined_coord.bed` contains coordinates.

```{sh}
head hg19.cage_peak_phase1and2combined_coord.bed
```

I want to replace the names in the coordinates file (fourth field) with the
`short_description` from the first files.  But the files are not sorted in the
same order.  Let's do the work in `R`

## Merge and output in `R`

```{r}
ann <- read.table( "hg19.cage_peak_phase1and2combined_ann.txt", head = TRUE
                 , sep = "\t", row.names = 1)
summary(ann)
head(ann)
```

```{r}
coord <- read.table( "hg19.cage_peak_phase1and2combined_coord.bed"
                   , head = FALSE
                   , sep = "\t"
                   , row.names = 4)
colnames(coord) <- c("chrom", "start", "end", "score", "strand", "bstart", "bend", "col")
summary(coord)
head(coord)
```

```{r}
out <- merge(ann, coord, by = "row.names")[ ,c("chrom", "start","end","short_description","score","strand")]
out <- out[order(out$chrom, out$start, out$end, out$strand),]
summary(out)
head(out)
```

```{r}
write.table( out
           , file      = "hg19.cage_peak_phase1and2combined_anncoord.bed"
           , sep       = "\t"
           , row.names = FALSE
           , col.names = FALSE
           , quote     = FALSE)
```

## Check the output in the command line

```{sh}
wc -l hg19.cage_peak_phase1and2combined*bed
```

```{sh}
head hg19.cage_peak_phase1and2combined_anncoord.bed
```

