---
title: "Annotated DPI clusters in BED format."
output:
  html_document: 
    keep_md: yes
  html_notebook: default
---

## Goal:

Transform this:

```
chr1	564571	564600	chr1:564571..564600,+	2398	+	564588	564589	255,0,0
chr1	564639	564649	chr1:564639..564649,+	220	+	564645	564646	255,0,0
chr1	565266	565278	chr1:565266..565278,+	535	+	565269	565270	255,0,0
chr1	565478	565483	chr1:565478..565483,+	106	+	565480	565481	255,0,0
chr1	565509	565541	chr1:565509..565541,+	3594	+	565523	565524	255,0,0
chr1	565662	565697	chr1:565662..565697,+	3243	+	565695	565696	255,0,0
chr1	565857	565899	chr1:565857..565899,+	592	+	565862	565863	255,0,0
chr1	566537	566573	chr1:566537..566573,+	306	+	566564	566565	255,0,0
chr1	566736	566770	chr1:566736..566770,+	1145	+	566760	566761	255,0,0
chr1	566871	566878	chr1:566871..566878,+	102	+	566874	566875	255,0,0
```

Into that:

```
chr1	564571	564600	p1@MTND1P23	2398	+
chr1	564639	564649	p3@MTND1P23	220	+
chr1	565266	565278	p3@MTND2P28	535	+
chr1	565478	565483	p4@MTND2P28	106	+
chr1	565509	565541	p1@MTND2P28	3594	+
chr1	565662	565697	p@chr1:565662..565697,+	3243	+
chr1	565857	565899	p@chr1:565857..565899,+	592	+
chr1	566537	566573	p4@uc001aaz.2	306	+
chr1	566736	566770	p2@uc001aaz.2	1145	+
chr1	566871	566878	p7@uc001aaz.2	102	+
```

[FANTOM5](http://fantom.gsc.riken.jp/5/) CAGE peaks
([DPI clusters](http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/))
are avaliable as BED files for their coordinates, supplemented by an annotation
file.  This annotation files provides a short description that contains the
gene symbol when possible.  Here, I replace the "name" field of the BED file
by the "short_description" field of the annotation file.  I do it in R because
the files are not sorted in the same order, and I am more comfortable
with join operations in R than in shell.

## Get and check the data in the command line:

Download the human DPI clusters from the FANTOM5 website.

```{sh}
curl --silent http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/hg19.cage_peak_phase1and2combined_ann.txt.gz | gunzip > hg19.cage_peak_phase1and2combined_ann.txt
curl --silent http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/hg19.cage_peak_phase1and2combined_coord.bed.gz | gunzip > hg19.cage_peak_phase1and2combined_coord.bed
```

The file `hg19.cage_peak_phase1and2combined_ann.txt` contains annotations.

```{sh}
head hg19.cage_peak_phase1and2combined_ann.txt
```

The files `hg19.cage_peak_phase1and2combined_coord.bed` contains coordinates.

```{sh}
head hg19.cage_peak_phase1and2combined_coord.bed
```

I want to replace the names in the coordinates file (fourth field) with the
`short_description` from the first files.  But the files are not sorted in the
same order.  Let's do the work in `R`

## Merge and output in `R`

```{r}
ann <- read.table( "hg19.cage_peak_phase1and2combined_ann.txt", head = TRUE
                 , sep = "\t", row.names = 1)
summary(ann)
head(ann)
```

```{r}
coord <- read.table( "hg19.cage_peak_phase1and2combined_coord.bed"
                   , head = FALSE
                   , sep = "\t"
                   , row.names = 4)
colnames(coord) <- c("chrom", "start", "end", "score", "strand", "bstart", "bend", "col")
summary(coord)
head(coord)
```

```{r}
out <- merge(ann, coord, by = "row.names")[ ,c("chrom", "start","end","short_description","score","strand")]
out <- out[order(out$chrom, out$start, out$end, out$strand),]
summary(out)
head(out)
```

```{r}
write.table( out
           , file      = "hg19.cage_peak_phase1and2combined_anncoord.bed"
           , sep       = "\t"
           , row.names = FALSE
           , col.names = FALSE
           , quote     = FALSE)
```

## Check the output in the command line

```{sh}
wc -l hg19.cage_peak_phase1and2combined*bed
```

```{sh}
head hg19.cage_peak_phase1and2combined_anncoord.bed
```


## Same for mouse

```{sh}
curl --silent http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/mm9.cage_peak_phase1and2combined_ann.txt.gz | gunzip > mm9.cage_peak_phase1and2combined_ann.txt
curl --silent http://fantom.gsc.riken.jp/5/datafiles/latest/extra/CAGE_peaks/mm9.cage_peak_phase1and2combined_coord.bed.gz | gunzip > hg19.cage_peak_phase1and2combined_coord.bed
```

```{r}
ann   <- read.table( "mm9.cage_peak_phase1and2combined_ann.txt", head = TRUE
                   , sep = "\t", row.names = 1)

coord <- read.table( "hg19.cage_peak_phase1and2combined_coord.bed"
                   , head = FALSE
                   , sep = "\t"
                   , row.names = 4)
colnames(coord) <- c("chrom", "start", "end", "score", "strand", "bstart", "bend", "col")

out <- merge( ann, coord
            , by = "row.names")[ ,c("chrom", "start","end","short_description","score","strand")]
out <- out[order(out$chrom, out$start, out$end, out$strand),]

write.table( out
           , file      = "mm9.cage_peak_phase1and2combined_anncoord.bed"
           , sep       = "\t"
           , row.names = FALSE
           , col.names = FALSE
           , quote     = FALSE)
```

```{sh}
wc -l mm9.cage_peak_phase1and2combined*bed
head mm9.cage_peak_phase1and2combined_anncoord.bed
```


## Session information 

```{r sessionInfo}
sessionInfo()
```