```{r echo=FALSE}
options(width=100)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_knit$set(verbose = TRUE)
```

Simple interrogation of FANTOM5's SDRF files
============================================

Some of the FANTOM5 metadata is distributed in the _Sample and Data
Relationship Format_ ([SDRF](https://en.wikipedia.org/wiki/SDRF)) format along
with the alignment files in <http://fantom.gsc.riken.jp/5/datafiles/>.  Each
line describes one FANTOM5 library, except the first line that is a header.

Here is one example downloaded from the FANTOM5 website.

```{r sdrf_Example_Full, engine="sh", comment=NA}
curl http://fantom.gsc.riken.jp/5/datafiles/phase1.2/basic/human.cell_line.LQhCAGE/00_human.cell_line.LQhCAGE.hg19.assay_sdrf.txt
```

As you see there are long lines with multiple fields.  We will use the
3<sup>rd</sup> (`Comment [sample_name]`) and the 14<sup>th</sup> (`Library
Name`).  Here is their contents in the above example.

```{r sdrf_Example_Fields, engine="sh", comment=NA}
curl http://fantom.gsc.riken.jp/5/datafiles/phase1.2/basic/human.cell_line.LQhCAGE/00_human.cell_line.LQhCAGE.hg19.assay_sdrf.txt | cut -f 3,14
``` 

Simple shell functions to return or query library IDs
-----------------------------------------------------

The file `sdrf.sh` distributed here defines two simple shell functions,
`SDRFlib` and `SDRFdesc`.  These functions expect the FANTOM5 SDRF files to be
in the current directory or in a directory indicated by the environment
variable `F5`.  In the examples later, we load these functions with the command
`source sdrf.sh`.

```{r, sdrf_show_functions, engine="sh", cache=FALSE, comment=NA}
cat sdrf.sh
```

Download the SDRF files
-----------------------

The FANTOM5 data is organised in multiple subdirectories, each containing a
single SDRF file.  Here, we will download all the SDRF files in a single
directory.

```{r sdrf_lftp, engine="sh"}
PHASE='phase1.2'
echo 'mget */*sdrf.txt' |
  lftp http://fantom.gsc.riken.jp/5/datafiles/$PHASE/basic
wc -l *sdrf.txt
```

_Note: the SDRF files have one header line, so the total number of libraries is
not the total number of lines._

Library names (column 14) start with `CNhs` for HeliScopeCAGE libraries.  Phase
1.2 only contains HeliScopeCAGE libraries.

```{r sdrf_count_libraries, engine="sh"}
cut -f 14 *sdrf.txt |
  grep -v 'Library Name' |   # Remove header lines.
  cut -c 1-4 |
  sort | uniq -c | sort -n
```

Examples of use
---------------

Find all libraries with _pancreas_ in their description.

```{r sdrf_SDRFlib_example, engine="bash"}
source sdrf.sh

SDRFlib pancreas
```

Find the library name of all the libraries with _pancreas_ in their
description.

```{r sdrf_SDRFdesc_example, engine="bash"}
source sdrf.sh

for lib in $(SDRFlib pancreas);
do
  echo -ne "$lib\t"
  SDRFdesc $lib
done
```

Conclusion
----------

We will use these functions to get list the name of all human and mouse FANTOM5
libraries in the next tutorial.

Note also that there is more metadata and more expressive ways to interrogate
it, see <http://fantom.gsc.riken.jp/views/> for details.
