```{r echo=FALSE}
options(width=100)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_knit$set(verbose = TRUE)
```

Simplistic interrogation of FANTOM5's SDRF files
================================================

Simple shell functions to return or query library IDs
-----------------------------------------------------

should also work with $F5 pointed to the full mirror ?

```{r, sdrf_show_functions, engine="sh", cache=FALSE, comment=NA}
cat sdrf.sh
```

Download the SDRF files
-----------------------

```{r sdrf_lftp, engine="sh"}
PHASE='phase1.2'
echo 'mget */*sdrf.txt' |
  lftp http://fantom.gsc.riken.jp/5/datafiles/$PHASE/basic
wc -l *sdrf.txt
```

_Note: the SDRF files have one header line, so the total number of libraries is not the total number of lines._

Library names are in column 14 and start with `CNhs` for HeliScopeCAGE
libraries.  Phase 1.2 only contains HeliScopeCAGE libraries.

```{r sdrf_count_libraries, engine="sh"}
cut -f 14 *sdrf.txt |
  grep -v 'Library Name' |   # Remove header lines.
  cut -c 1-4 |
  sort | uniq -c | sort -n
```

Examples of use
---------------

Find example that also return sRNA libraries

```{r sdrf_SDRFlib_example, engine="bash"}
source sdrf.sh

SDRFlib pancreas
```

```{r sdrf_SDRFdesc_example, engine="bash"}
source sdrf.sh

for lib in $(SDRFlib pancreas);
do
  echo -ne "$lib\t"
  SDRFdesc $lib
done
```

