```{r echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
options(width=150)
```

Digital expression comparison between nanoCAGE libraries testing variant template-switching oligonucleotides
============================================================================================================

This tutorial is based on the supplementary material of the manuscript
_Comparison of RNA- and LNA-Hybrid Oligonucleotides in Template-Switching
Reactions_ ([Harbers et al.,2013](http://dx.doi.org/10.1186/1471-2164-14-665)),
that documented the commands run to compare the expression levels of CAGE
clusters in libraries made with different template-switching oligonucleotides.
It is an example comparison of shallow-sequenced nanoCAGE libraries.

See the main [README](../README.md) for general recommendations on how or what
to prepare before running this tutorial.

Table of contents
-----------------

 * [Data download and preparation](#data-prep)
 * [Artifact cleaning and alignment of the reads](#artifact-cleaning-and-alignment)
 * [Tag clustering](#clustering)
 * [Annotation](#annotation)
 * [Preparation for statistical analysis](#r-prep)
 * [Differential representation analysis](#differential-expression)
   * [LNA vs. RNA](#LNA-RNA)
   * [RNA vs. DNA](#RNA-DNA)
   * [DNA vs. LNA](#DNA-LNA)
 * [Output of the results as tables](#make-tables)
 * [Notes on the software](#notes-on-software)


<a id="data-prep">Data download and preparation</a>
---------------------------------------------------

### Information and download

The `DRR014141` library (`NCms10010` in the original publication) was made with
total RNA from rat muscle.  It is comparing template-switching oligonucleotides
that end in RNA (r), DNA (d) or LNA (l) bases.  The comparisons were
multiplexed in triplicates.

The data is a single-end MiSeq run (ID: `121012_M00528_0022_AMS2003997-00050`)
of 4,682,200 reads.  The following commands download (`wget`) the data from
[DRA](http://trace.ddbj.nig.ac.jp/dra/index.html) (DDBJ's Sequence Read Archive)
and test its integrity (`md5sum`).

```{r download_material, engine="bash"}
wget --quiet --timestamping ftp://ftp.ddbj.nig.ac.jp/ddbj_database/dra/fastq/DRA001/DRA001167/DRX012672/DRR014141.fastq.bz2
echo "b0b3070fbbae0073507234d9048325ff  DRR014141.fastq.bz2" | md5sum -c
```

### Sample names and barcodes

The following command produces a file called `DRR014141.id` that associates the
barcodes and sample identifiers (_IDs_).  The sample IDs encode the type of
template-switching oligonucleotide used, where each letter representing the
chemical nature of the third, second and first nucleotide, from the 3′ end.
This file will be used to rename the samples after demultiplexing, and to
iterate over the samples in some steps of the processing.

```{r create_ID_file, engine="bash"}
cat <<__ID__ > DRR014141.id
rrr_1	CACTGA
rrr_2	GCTCTC
rrr_3	TCGCGT
ddd_1	ATCGTG
ddd_2	CACGAT
ddd_3	GTATAC
ddl_1	ACAGAT
ddl_2	CTGACG
ddl_3	GAGTGA
dll_1	AGTAGC
dll_2	GCTGCA
dll_3	TCGAGC
lll_1	ATCATA
lll_2	CGATGA
lll_3	TATAGC
__ID__
```

<a id="demultiplex-trim-filter">Demultiplexing, trimming and filtering</a>
--------------------------------------------------------------------------

Samples are demultiplexed with [TagDust](http://bioinformatics.oxfordjournals.org/content/25/21/2839)
version 2.  One key feature of TagDust it has a declarative syntax that enables
automatic trimming of the linkers and at the same times leaves a written record
of the read structure.

Download [TagDust 2](http://sourceforge.net/projects/tagdust/) and install it
in the user's path.  See the main [README](../README.md) for details.

### Removal of artifacts with TagDust

The following command creates a small FASTA file recording the sequence of
possible sources of artefact reads.

```{r tagdust, engine="bash"}
cat > tagdust.fa <<__TagDust__
>TS (before barcode)
TAGTCGAACTGAAGGTCTCCAGCA
>RT (without random bases)
TAGTCGAACTGAAGGTCTCCGAACCGCTCTTCCGATCT
>empty (TS linker + RT reverse-complemented)
TATAGGGAGATCGGAAGAGCGGTTCGGAGACCTTCAGTTCGACTA
__TagDust__
```

### Read structure

In this version of nanoCAGE, the reads start with 6 bases of barcode, 8 bases
of random fingerprint, 4 bases of spacer, and 3 base of linker.  In the syntax
of TagDust, this translates as follows:

```
-1 O:N
-2 B:CACTGA,GCTCTC,TCGCGT,ATCGTG,CACGAT,GTATAC,ACAGAT,CTGACG,GAGTGA,AGTAGC,GCTGCA,TCGAGC,ATCATA,CGATGA,TATAGC
-3 F:NNNNNNNN
-4 S:TATAGGG
-5 R:N
```

To avoid manual errors, the list of barcodes can be generated with the
following command, that takes the second column of the file `DRR014141.id`,
replace return characters by commas, and then removes the final comma.  The
`echo` command is here to ease the display of the result.

```
awk '{print $2}' DRR014141.id | perl -pe 's/\n/,/' | sed 's/,$//' ; echo
```

### Run TagDust

This will take time.

```
BARCODES=$(awk '{print $2}' DRR014141.id | perl -pe 's/\n/,/' | sed 's/,$//')
tagdust -t8 -ref tagdust.fa -1 O:N -2 B:$BARCODES -3 F:NNNNNNNN -4 S:TATAGGG -5 R:N -l . -o DRR014141 DRR014141.fastq.bz2
```


### Note on FASTX-toolkit

Before TagDust 2 was written, we were using [FASTX-toolkit](http://hannonlab.cshl.edu/fastx_toolkit/).
For the record, here is an example of command.

```
bzcat DRR014141.fastq.bz2 fastx_barcode_splitter.pl --bcfile DRR014141.id --prefix DRR014141. --suffix .fq --bol --exact
```

The log output can be converted to a tab-separated format by piping it to shell commands like the following.

```
sed 1d | cut -f1,2 |
  perl -ne 'print "extracted\t$_"' |
  grep -v -e unmatched -e total
```

Assuming that there are no indels, the linker sequences in the reads
demultiplexed with `fastx_barcode_splitter.pl` can then be removed with a
command such as the following.  (In version `0.0.14` or higher of FASTX
Toolkit, the `-Q33` option is not necessary anymore).

```
for FASTQ in *.fq
do
  fastx_trimmer -f 22 -Q33 < $FASTQ | sponge $FASTQ
done
```

The `sponge` command is from the [moreutils](http://joeyh.name/code/moreutils/)
collection.  See the main [README](../README.md) for details.

Trimmed reads would then be screened for artefacts with TagDust 1 using the
same definition as above.


### Note on _log_ files.

In this tutorial, the output of some commands is captured with [Unix
pipes](Pipeline_(Unix)), and transformed into text files where each line is a
numerical summary (number of reads extracted, filtered, aligned, etc) for a
given sample, for instance, `extracted	ddd_1	333255`.  The `tee` command
prints the results while the file is saved.


### Alignment on the rat genome version 4

The following assumes the genome downloaded and indexed for BWA in the current
directory, using `rn4_male` as a base name (note that the base name is defined
by the `-p` option in the command `bwa index`, and that in a lot of
facility-provided or internet-provided indexes (such as the Illumina
[iGenomes](http://cufflinks.cbcb.umd.edu/igenomes.html)), the path to give to
BWA ends in `.fa`.

```{r align, engine="bash", dependson='tagdust'}
GENOME=rn4_male
for FQ in *.dusted.fastq
do
  bwa aln -t8 $GENOME -f $(basename $FQ .dusted.fastq).sai $FQ
  bwa samse   $GENOME    $(basename $FQ .dusted.fastq).sai $FQ |
    samtools view -uS - |
    samtools sort - $(basename $FQ .dusted.fastq)
done
```

### Filter reads aligning to rDNA with rRNAdust

`rRNAdust` is available in the supplementary material at <http://genome.gsc.riken.jp/plessy-20130430/rRNAdust_1.02.tar.gz>.

Download the reference rRNA sequences at <http://www.ncbi.nlm.nih.gov/nuccore/V01270.1> and <http://www.ncbi.nlm.nih.gov/gene/170603>, and save them in a file called `rat_rDNA.fa`.

The reason for removing reads matching rRNA sequences is that in most reference
genomes, the nuclear rRNA genes are not included because they are repeated and
hard to assemble.  So if the reads originating from rRNA are not filtered out,
they will align at suboptimal places in the genome that resemble the rRNA, and
create artificial signal.  In addition, these artefacts can also make the
alignment files considerably larger, because of the propensy of rRNA sequences
to match multiple genomic regions, in particular rRNA-derived repeated elements.

```{r rrnadust, engine="bash", dependson='align'}
for ID in $( awk '{print $1}' DRR014141.id )
do
  echo -ne "rdna\t$ID\t"
  (rRNAdust -t8 rat_rDNA.fa DRR014141.$ID.bam | samtools view -bS  - 2> /dev/null | sponge DRR014141.$ID.bam) 2>&1 | sed 's/Excluded: //'
done | tee DRR014141.rdna.log
```

Note that _rRNAdust_ has been integrated in the version 2 of _TagDust_,
available at <http://tagdust.sourceforge.net/>.  This tutorial will be updated
later accordingly.

### Alignment statistics

```{r align_stats, engine="bash", dependson='rrnadust'}
for ID in $( awk '{print $1}' DRR014141.id )
do
  echo -ne "mapped\t$ID\t"
  (samtools flagstat DRR014141.$ID.bam | grep mapped | grep %) | cut -f1 -d' '
done | tee DRR014141.mapped.log
```

### Filter out possible strand-invasion tags

Strand invasion was described in [Tang et al, Nucl. Acids Res. (2013) 41
(3):e44](http://nar.oxfordjournals.org/content/41/3/e44), and is more frequent
in LNA- or DNA-based template-switching oligonucleotides.

The following commands need an updated version of the `find_strand_invasion.pl`
that was in Tang et al's supplementary material, where a new `-f` option is
added, with the same semantics as in `samtools`.  It is available at
<https://github.com/davetang/23180801>.


```{r strand_invasion, engine="bash", dependson='rrnadust'}
ERRORS=2
for BAM in DRR014141.???_?.bam
do
  find_strand_invasion.pl -f $BAM -g rn4_male.fa -e $ERRORS -s TATA
done
for ID in $( awk '{print $1}' DRR014141.id )
do
  echo -ne "strand-invasion-$ERRORS\t$ID\t"
  (samtools flagstat DRR014141.${ID}_nw_${ERRORS}_??????_removed_sorted.bam | grep mapped | grep %) | cut -f1 -d' '
done | tee DRR014141.strand-invasion-$ERRORS.log
```

<a name="clustering">Tag clustering</a>
---------------------------------------

Level 1 clusters are single-nucleotide resolution data representing the 5′ ends
of the CAGE tags.  Level 2 clusters are groups of level 1 clusters that are
separated by 20 or less nucleotides.

The `level1.py` and `level2.py` scripts implement tag clustering like in the
[FANTOM3](http://fantom.gsc.riken.jp/3/) and
[FANTOM4](http://fantom.gsc.riken.jp/4/) projects, and are available at
<http://genome.gsc.riken.jp/plessy-20130430/PromoterPipeline_20130430.tar.gz>.

They output their results in Order Switchable Column
([OSC](http://sourceforge.net/projects/osctf/)) format, where each line is a
cluster, and each library gives one column counting the tags in the clusters,
and another column where the counts are normalised in parts per million.

See also [Carninci et al., Nature Genetics 38
626-635 (2006)](http://www.nature.com/ng/journal/v38/n6/full/ng1789.html) and
[Suzuki et al., Nature Genetics 41 553-562
(2009)](http://www.nature.com/ng/journal/v41/n5/full/ng.375.html) for original
examples of CAGE tag clustering.

```{r tag_clustering, engine="bash", dependson='strand_invasion'}
level1.py -o DRR014141.l1.osc.gz -F 516 \
  DRR014141.???_?.bam \
  DRR014141.???_?_nw_?_??????_filtered_sorted.bam

level2.py -o DRR014141.l2.osc -t 0 DRR014141.l1.osc.gz
gzip DRR014141.l2.osc
```

The resulting file `DRR014141.l1.osc.gz` can be loaded in the
[Zenbu](http://fantom.gsc.riken.jp/zenbu/) system to browse the alignments on
the rat genome.

<a name="annotation">Annotation</a>
-----------------------------------

### Preparation of the reference files.

Data from ENSEMBL 69 were retrieved via
[BioMart](http://oct2012.archive.ensembl.org/biomart/martview/), with the following XML query.
Note that `external_gene_id` is called _Associated Gene Name_ in the web
interface.

```
<?xml version="1.0" encoding="UTF-8"?>
<!DOCTYPE Query>
<Query  virtualSchemaName = "default" formatter = "TSV" header = "0" uniqueRows = "0" count = "" datasetConfigVersion = "0.6" >
			
	<Dataset name = "rnorvegicus_gene_ensembl" interface = "default" >
		<Attribute name = "ensembl_gene_id" />
		<Attribute name = "ensembl_transcript_id" />
		<Attribute name = "chromosome_name" />
		<Attribute name = "strand" />
		<Attribute name = "external_gene_id" />
		<Attribute name = "transcript_start" />
		<Attribute name = "transcript_end" />
		<Attribute name = "gene_biotype" />
		<Attribute name = "exon_chrom_start" />
		<Attribute name = "exon_chrom_end" />
	</Dataset>
</Query>
```

Convert coordinates to gene names

```{r convert_biomart, engine="bash"}
cat mart_export.txt | 
  sed -e 1d -e 's|\t\t|\tno_symbol\t|'           |
  awk '{OFS="\t"} {print $3, $6, $7, $5, 0, $4}' |
  grep -v ^[JA]                                  |
  uniq                                           |
  sed -e 's/-1$/-/'                              \
      -e 's/1$/+/'                               \
      -e 's/^/chr/'                              |
  sort -k1,1 -k2,2n                              \
  > rn4_male.symbols.bed


### Gene symbols

```{r gene_symbols, engine="bash", dependson='convert_biomart'}
zcat DRR014141.l1.osc.gz | grep -v \# | sed 1d | awk '{OFS="\t"}{print $2, $3, $4, "l2", "1000", $5}' > DRR014141.l1.bed
zcat DRR014141.l2.osc.gz | grep -v \# | sed 1d | awk '{OFS="\t"}{print $2, $3, $4, "l2", "1000", $5}' > DRR014141.l2.bed
zcat DRR014141.l2.osc.gz | grep -v \# | sed 1d | awk '{OFS="\t"}{print $2, $3, $4, "l2", "1000", $5}' > DRR014141.l2.bed

bedtools intersect -a DRR014141.l2.bed -b rn4_male.symbols.bed -s -loj |
  awk '{OFS="\t"}{print $1":"$2"-"$3$6,$10}' | 
  bedtools groupby -g 1 -c 2 -o distinct > DRR014141.l2.genes
```

### Repeated elements

Download the repeatmasker track from the [UCSC genome browser](http://genome.ucsc.edu) and save it in a file called `rn4_male.repeatmasker.bed`.

```{r intersect_rmsk, engine="bash"}
bedtools intersect -a DRR014141.l2.bed -b rn4_male.repeatmasker.bed -s -loj |
  awk '{OFS="\t"}{print $1":"$2"-"$3$6,$10}' |
  bedtools groupby -g 1 -c 2 -o distinct > DRR014141.l2.rmsk
```

<a name="r-prep">Preparation for statistical analysis</a>
---------------------------------------------------------

The following commands are run in the [`R` package for statistical computing](http://www.r-project.org/).

### Load the data.

The oscR library is available at <https://github.com/charles-plessy/oscR>.
Note that OSCTables can also be loaded with R's `read.table()` command (but pay
attention to comment and quote characters, etc.).

The following commands load the level 1 and 2 clusters into data frames where
the column names correspond to the sample IDs defined above.

```{r load_data, dependson='tag_clustering'}
library(oscR)

l1 <- read.osc("DRR014141.l1.osc.gz", drop.coord=T, drop.norm=T)
l2 <- read.osc("DRR014141.l2.osc.gz", drop.coord=T, drop.norm=T)

colnames(l1) <- sub('raw.DRR014141.','',colnames(l1))
colnames(l2) <- sub('raw.DRR014141.','',colnames(l2))

colnames(l1) <- sub('_......_filtered_sorted', '', colnames(l1))
colnames(l2) <- sub('_......_filtered_sorted', '', colnames(l2))
```

### Organise the data.

The following commands defined convenient shortcuts to manipulates groups of
libraries. The presence of `nw_2` in the names indicate that strand-invasion
artifacts have been removed.

```{r define_aliases}
ddd <- c('ddd_1', 'ddd_2', 'ddd_3')
ddl <- c('ddl_1', 'ddl_2', 'ddl_3')
dll <- c('dll_1', 'dll_2', 'dll_3')
lll <- c('lll_1', 'lll_2', 'lll_3')
rrr <- c('rrr_1', 'rrr_2', 'rrr_3')
all <- c(rrr, lll, dll, ddl, ddd)

ddd_nw_2 <- c('ddd_1_nw_2', 'ddd_2_nw_2', 'ddd_3_nw_2')
ddl_nw_2 <- c('ddl_1_nw_2', 'ddl_2_nw_2', 'ddl_3_nw_2')
dll_nw_2 <- c('dll_1_nw_2', 'dll_2_nw_2', 'dll_3_nw_2')
lll_nw_2 <- c('lll_1_nw_2', 'lll_2_nw_2', 'lll_3_nw_2')
rrr_nw_2 <- c('rrr_1_nw_2', 'rrr_2_nw_2', 'rrr_3_nw_2')
all_nw_2 <- c(rrr_nw_2, lll_nw_2, dll_nw_2, ddl_nw_2, ddd_nw_2)
```

```{r calculate_means_and_SDs, dependson='load_data'}
TPM <- function(clusters){
  clusters.tpm <- data.frame(prop.table(as.matrix(clusters),2) * 1000000)
  colnames(clusters.tpm) <- colnames(clusters)
  return(clusters.tpm)
}

L2 <- TPM(l2)

L2.means <- data.frame(
  ddd = apply(L2[,ddd],1,mean),
  ddl = apply(L2[,ddl],1,mean),
  dll = apply(L2[,dll],1,mean),
  lll = apply(L2[,lll],1,mean),
  rrr = apply(L2[,rrr],1,mean) )

L2.means_nw_2 <- data.frame(
  ddd = apply(L2[,ddd_nw_2],1,mean),
  ddl = apply(L2[,ddl_nw_2],1,mean),
  dll = apply(L2[,dll_nw_2],1,mean),
  lll = apply(L2[,lll_nw_2],1,mean),
  rrr = apply(L2[,rrr_nw_2],1,mean) )

L2.sd <- data.frame(
  ddd = apply(L2[,ddd],1,sd),
  ddl = apply(L2[,ddl],1,sd),
  dll = apply(L2[,dll],1,sd),
  lll = apply(L2[,lll],1,sd),
  rrr = apply(L2[,rrr],1,sd) )
```

### Annotation of the results

```{r load_gene_symbols, dependson=c('gene_symbols','intersect_rmsk')}
genesymbols <- read.table("DRR014141.l2.genes", col.names=c("cluster","symbol"))
rownames(genesymbols) <- genesymbols$cluster
genesymbols$rmsk <- read.table("DRR014141.l2.rmsk", col.names=c("cluster","rmsk"))[,'rmsk']
```

<a name="differential-expression">Differential representation analysis<a>
-------------------------------------------------------------------------

Statistical comparisons using [edgeR](http://www.bioconductor.org/packages/release/bioc/html/edgeR.html).

```{r load_edgeR, message=FALSE, dependson='load_data'}
library(edgeR)
```

The following plots represent:

 * the multidimensional scaling of the samples,
   where spatial separation between the two sets of triplicates indicates that the
   factor that is compared (type of template-switching oligonucleotide) introduces
   more differences that the simple technical fluctuations,

 * the expression levels of the CAGE clusters as a M-A plot, where dots in red are clusters
   significantly enriched in one type of libraries.  Vertical distance from the
   horizontal midline represent the amplitude of the differences, and distance on
   the horizontal axis represents the average strength of expression.

The following comparisons show the difference (or lack of it) between
non-filtered and filtered data, and then explore the filtered data in more
details.

Note on row names: older versions of edgeR will return `NULL` with commands
such as `rownames(lr.com)`, where `lr.com` is an object returned by the
`exactTest` command.  As a result, the plots will lack the red-color on the
differentially expressed clusters.  The solution is to upgrade edgeR.

### <a name='LNA-RNA'>LNA vs. RNA</a>

```{r LNA_vs_RNA, dev='png'}
x     <- DGEList(counts=l2[,c(lll,rrr)], group=c(rep('lll',3),rep('rrr',3)), remove.zeros=TRUE)
x     <- calcNormFactors(x)
x     <- estimateCommonDisp(x)
x     <- estimateTagwiseDisp(x)
x.com <- exactTest(x)
lr <- x
lr.com <- x.com
plotMDS(lr)
plotSmear(lr.com, de.tags=rownames(lr.com)[decideTestsDGE(lr.com) != 0], cex=0.8, main='LNA / RNA', ylab="LNA (bottom) / RNA (top)")

x     <- DGEList(counts=l2[,c(lll_nw_2,rrr_nw_2)], group=c(rep('lll_nw_2',3),rep('rrr_nw_2',3)), remove.zeros=TRUE)
x     <- calcNormFactors(x)
x     <- estimateCommonDisp(x)
x     <- estimateTagwiseDisp(x)
x.com <- exactTest(x)
lr_nw_2 <- x
lr_nw_2.com <- x.com
plotMDS(lr_nw_2)
plotSmear(lr_nw_2.com, de.tags=rownames(lr_nw_2.com)[decideTestsDGE(lr_nw_2.com) != 0], cex=0.8, main='LNA / RNA (filtered)', ylab="LNA (bottom) / RNA (top)")
```

```{r compare_LNA_RNA_numbers, results="hide"}
lr_nw_2.up      <- sum(decideTestsDGE(lr_nw_2.com) > 0)
lr_nw_2.down    <- sum(decideTestsDGE(lr_nw_2.com) < 0)
```

`r lr_nw_2.up` clusters were enriched and `r lr_nw_2.down` were depleted in RNA
libraries compared to LNA.  The top 100 stronger fold changes in each direction are shown below.

```{r compare_LNA_RNA, dependson=c('load_gene_symbols' ,'LNA_vs_RNA,'), echo=-1}
options(width=150)
# Summary of the top 100 clusters enriched in RNA libraries.
summary(merge(subset(topTags(lr_nw_2.com, Inf)$table, logFC > 0)[1:100,], genesymbols[,-1], by=0, sort=FALSE))
# Top 15 clusters enriched in RNA libraries.
merge(subset(topTags(lr_nw_2.com, Inf)$table, logFC > 0)[1:15,], genesymbols[,-1], by=0, sort=FALSE)
# Summary of the top 100 clusters enriched in LNA libraries.
summary(merge(subset(topTags(lr_nw_2.com, Inf)$table, logFC < 0)[1:100,], genesymbols[,-1], by=0, sort=FALSE))
```

Positive fold change indicate enrichment in RNA libraries.  The 7SLRNA hits
are concentrated at the top of the list.  LNA libraries are enriched for hits
on CAGAGA repeats, even beyond the significance level (FDR) of the statistical
comparison.

```{r srprna, dependson='load_gene_symbols', dev='svg'}
library(reshape)
library(ggplot2)
srprna <- rownames(subset(genesymbols, rmsk=="7SLRNA"))
srprna.expression <- melt(L2[srprna, c(rrr_nw_2, ddd_nw_2, lll_nw_2)], measure.vars=c(rrr_nw_2, ddd_nw_2, lll_nw_2))
for (group in c('ddd','lll','rrr')) srprna.expression[grep(group,srprna.expression$variable),'group'] <- group
srprna.expression$group <- reorder(factor(srprna.expression$group), srprna.expression$value, function(x) sum(x) * -1)
qplot(data=srprna.expression, value, reorder(variable, value, sum), xlab="Parts per million", ylab="Library", main="Expression levels of 7SL RNA genes", col=group)
```

The measured expression of 7SL RNA is strongest in RNA libraries, strong in DNA
libraries, and weak in LNA libraries.

### <a name='RNA-DNA'>RNA vs. DNA</a>

```{r RNA_vs_DNA, dev='png'}
x     <- DGEList(counts=l2[,c(rrr,ddd)], group=factor(c(rep("rrr", 3), rep("ddd", 3)), levels=c("rrr", "ddd")), remove.zeros=TRUE)
x     <- calcNormFactors(x)
x     <- estimateCommonDisp(x)
x     <- estimateTagwiseDisp(x)
x.com <- exactTest(x)
rd <- x
rd.com <- x.com
plotMDS(x)
plotSmear(rd.com, de.tags=rownames(rd.com)[decideTestsDGE(rd.com) != 0], cex=0.8, main='RNA / DNA', ylab="RNA (bottom) / LNA (top)")

x     <- DGEList(counts=l2[,c(rrr_nw_2,ddd_nw_2)], group=factor(c(rep("rrr_nw_2", 3), rep("ddd_nw_2", 3)), levels=c("rrr_nw_2", "ddd_nw_2")), remove.zeros=TRUE)
x     <- calcNormFactors(x)
x     <- estimateCommonDisp(x)
x     <- estimateTagwiseDisp(x)
x.com <- exactTest(x)
plotMDS(x)
rd_nw_2 <- x
rd_nw_2.com <- x.com
plotSmear(rd_nw_2.com, de.tags=rownames(rd_nw_2.com)[decideTestsDGE(rd_nw_2.com) != 0], cex=0.8, main='RNA / DNA (filtered)', ylab="RNA (bottom) / DNA (top)")
```

```{r compare_RNA_DNA_numbers, echo=FALSE}
rd_nw_2.up      <- sum(decideTestsDGE(rd_nw_2.com) > 0)
rd_nw_2.down    <- sum(decideTestsDGE(rd_nw_2.com) < 0)
```

`r rd_nw_2.down` clusters were enriched and `r rd_nw_2.up` were depleted in RNA
libraries compared to DNA.

The majority of the clusters in the top 100 enriched in the RNA libraries did
not overlap with repeat elements, and were overlapping with loci having gene
symbols.  In contrast, the majority of the top 100 clusters enriched in the DNA
libraries did not overlap with known genes.  A mild enrichment for GGGTG
simple repeats is noted.

```{r compare_RNA_DNA, dependson=c('load_gene_symbols' ,'RNA_vs_DNA,')}
# Summary of the top 100 clusters enriched in DNA libraries.
summary(merge(subset(topTags(rd_nw_2.com, Inf)$table, logFC > 0)[1:100,], genesymbols[,-1], by=0, sort=FALSE))
# Summary of the top 100 clusters enriched in RNA libraries.
summary(merge(subset(topTags(rd_nw_2.com, Inf)$table, logFC < 0)[1:100,], genesymbols[,-1], by=0, sort=FALSE))
```

### <a name='DNA-LNA'>DNA vs. LNA</a>


```{r DNA_vs_LNA, dev='png'}
x     <- DGEList(counts=l2[,c(ddd,lll)], group=c(rep('ddd',3),rep('lll',3)), remove.zeros=TRUE)
x     <- calcNormFactors(x)
x     <- estimateCommonDisp(x)
x     <- estimateTagwiseDisp(x)
x.com <- exactTest(x)
dl <- x
dl.com <- x.com
plotMDS(x)
plotSmear(dl.com, de.tags=rownames(dl.com)[decideTestsDGE(dl.com) != 0], cex=0.8, main='DNA / LNA', ylab="DNA (bottom) / LNA (top)")

x     <- DGEList(counts=l2[,c(ddd_nw_2,lll_nw_2)], group=c(rep('ddd_nw_2',3),rep('lll_nw_2',3)), remove.zeros=TRUE)
x     <- calcNormFactors(x)
x     <- estimateCommonDisp(x)
x     <- estimateTagwiseDisp(x)
x.com <- exactTest(x)
plotMDS(x)
dl_nw_2 <- x
dl_nw_2.com <- x.com
plotSmear(dl_nw_2.com, de.tags=rownames(dl_nw_2.com)[decideTestsDGE(dl_nw_2.com) != 0], cex=0.8, main='DNA / LNA (filtered)', ylab="DNA (bottom) / LNA (top)")
```

```{r compare_DNA_LNA_numbers, echo=FALSE}
dl_nw_2.up      <- sum(decideTestsDGE(dl_nw_2.com) > 0)
dl_nw_2.down    <- sum(decideTestsDGE(dl_nw_2.com) < 0)
```

`r dl_nw_2.up` clusters were enriched and `r dl_nw_2.down` were depleted in LNA
libraries compared to DNA.

After filtering out strand-invasion artifacts, only few significant differences
remain between the DNA and LNA libraries.  7SLRNA was also depleted in LNA
libraries, and CAGAGA repeats were enriched.


```{r compare_DNA_LNA, dependson=c('load_gene_symbols' ,'DNA_vs_LNA,')}
summary(decideTestsDGE(dl_nw_2.com))
summary(merge(subset(topTags(dl_nw_2.com, Inf)$table, logFC > 0)[1:100,], genesymbols[,-1], by=0, sort=FALSE))
summary(merge(subset(topTags(dl_nw_2.com, Inf)$table, logFC < 0)[1:100,], genesymbols[,-1], by=0, sort=FALSE))
```

```{r save_workspace, cache=FALSE}
save.image('analysis.RData')
```

<a name="make-tables">Output of the results as tables</a>
--------------------------------------------------------

```{r make-tables, dependson=c('compare_LNA_RNA', 'compare_RNA_DNA', 'compare_DNA_LNA')}
# One table per list of significantly over-represented clusters.
write.csv(file='R-L.csv', merge(subset(topTags(lr_nw_2.com, Inf)$table, logFC > 0 & FDR < 0.1), genesymbols[, -1], by = "row.names", all.x='T', sort = FALSE), row.names=FALSE)
write.csv(file='L-R.csv', merge(subset(topTags(lr_nw_2.com, Inf)$table, logFC < 0 & FDR < 0.1), genesymbols[, -1], by = "row.names", all.x='T', sort = FALSE), row.names=FALSE)
write.csv(file='D-R.csv', merge(subset(topTags(rd_nw_2.com, Inf)$table, logFC > 0 & FDR < 0.1), genesymbols[, -1], by = "row.names", all.x='T', sort = FALSE), row.names=FALSE)
write.csv(file='R-D.csv', merge(subset(topTags(rd_nw_2.com, Inf)$table, logFC < 0 & FDR < 0.1), genesymbols[, -1], by = "row.names", all.x='T', sort = FALSE), row.names=FALSE)
write.csv(file='L-D.csv', merge(subset(topTags(dl_nw_2.com, Inf)$table, logFC > 0 & FDR < 0.1), genesymbols[, -1], by = "row.names", all.x='T', sort = FALSE), row.names=FALSE)
write.csv(file='D-L.csv', merge(subset(topTags(dl_nw_2.com, Inf)$table, logFC < 0 & FDR < 0.1), genesymbols[, -1], by = "row.names", all.x='T', sort = FALSE), row.names=FALSE)

# One summary table combining all the results.
l2[rownames(topTags(lr_nw_2.com, Inf)), "LR.logFC"] <- topTags(lr_nw_2.com, Inf)$table$logFC
l2[rownames(topTags(lr_nw_2.com, Inf)), "LR.FDR"]   <- topTags(lr_nw_2.com, Inf)$table$FDR
l2[rownames(topTags(rd_nw_2.com, Inf)), "RD.logFC"] <- topTags(rd_nw_2.com, Inf)$table$logFC
l2[rownames(topTags(rd_nw_2.com, Inf)), "RD.FDR"]   <- topTags(rd_nw_2.com, Inf)$table$FDR
l2[rownames(topTags(dl_nw_2.com, Inf)), "DL.logFC"] <- topTags(dl_nw_2.com, Inf)$table$logFC
l2[rownames(topTags(dl_nw_2.com, Inf)), "DL.FDR"]   <- topTags(dl_nw_2.com, Inf)$table$FDR
l2$symbol <- genesymbols$symbol
l2$rmsk <- genesymbols$rmsk
write.csv(file="DRR014141.DGE.csv", l2)
```


<a name="notes-on-software">Notes on the software</a>
-----------------------------------------------------

This analysis was done on a iMac with a i7 hyperthreaded quad-core CPU (2.93
GHz) and 12 GiB of memory, running [Debian](http://www.debian.org) system, with
the following packages installed.

```{r list-packages, engine="bash"}
dpkg -l bedtools bwa fastx-toolkit moreutils r-base r-bioc-edger samtools
```

This tutorial was made with the [knitr](http://yihui.name/knitr/) library for
`R`, that produces HTML pages from templates containing executable code.

```{r session_info, cache=FALSE}
sessionInfo()
```
