---
title: "FANTOM5 DPI peak shape"
author: "Charles Plessy"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

FANTOM5 promoter peak shape
===========================

The FANTOM5 expression atlas paper (Forrest et al., 2014) describes promoter
models produced by Decomposition-based Peak Identification (DPI) on CAGE data.
The DPI peaks have different widths according to the underlying CAGE signal;
the goal of this page is to use this width as a proxy to classify the promoters
in _sharp_ or _broad_ categories.

There are multiple limitations to this approach.  In particular, promoter shape
can change across tissues, and also across time in the same tissue (for instance
during development).  Indeed, the promoter shape analysis in Forrest 2014 (ext.
data fig. 2) was performed with a more robust approach, using the 0.1–0.9
interquantile width of the CAGE expression data accross all samples.  Moreover,
genes may have multiple promoters of different shapes.  Nevertheless, there are
cases, such as when handlings large groups of genes, where it may be useful to
estimate if these genes tend to have sharper or broader promoters.  This is the
aim of the approach presented here.


Width per DPI peak
------------------

First, run the example in `../FANTOM5_DPI_BED_file_with_annotation/`.

Then, load the result files.

```{r loadAnnotatedClusters}
(human <- rtracklayer::import.bed("../FANTOM5_DPI_BED_file_with_annotation/hg19.cage_peak_phase1and2combined_anncoord.bed"))
(mouse <- rtracklayer::import.bed("../FANTOM5_DPI_BED_file_with_annotation/mm9.cage_peak_phase1and2combined_anncoord.bed"))
```

Then, just use Bioconductor to calculate the width

```{r width}
human$width <- GenomicRanges::width(human)
mouse$width <- GenomicRanges::width(mouse)
```


Promoter shape per gene
------------------------

Here we attribute to a given gene symbol the properties if its promoter with the
highest expression, which by definition has a name starting with `p1`.


```{r perGene}
human <- human[substr(human$name, 1, 3) == "p1@"]
human$name  <- factor(sub("p1@", "", human$name))
human$name  <- factor(sub(",.*", "", human$name))
human

mouse <- mouse[substr(mouse$name, 1, 3) == "p1@"]
mouse$name  <- factor(sub("p1@", "", mouse$name))
mouse$name  <- factor(sub(",.*", "", mouse$name))
mouse
```


Save lists
----------

```{r saveLists}
write.table( file = "human_top_promoter_width.txt"
           , GenomicRanges::mcols(human)
           , row.names = FALSE, quote = FALSE, sep = "\t")

write.table( file = "mouse_top_promoter_width.txt"
           , GenomicRanges::mcols(mouse)
           , row.names = FALSE, quote = FALSE, sep = "\t")
```


Session information
-------------------

```{r sessionInfo}
sessionInfo()
```